---
title: "Lab 3b - Exercises"
author: "Environmental Data Analytics | John Fay and Luana Lima "
date: "Spring 2026"
output: pdf_document
---

## Set up the session
- Load packaged: tidyverse, lubridate, & here
- Import the `EPAair_PM25_NC2019_raw.csv` file
  - Set strings to be read in as factors
  - Set the "Site.ID" column to be read in as a factor
- Set the date column to be a date

```{r Setup}
#Import packages
library(tidyverse); library(lubridate); library(here)

#Check workspace
here()

#Import data
pm25 <- read.csv(
  here("Data","Raw","EPAair_PM25_NC2019_raw.csv"),
  #----ENTER THE FILENAME HERE, USING THE here() PACKAGE----
  stringsAsFactors = TRUE,
  colClasses = c("Site.ID"='factor')
)

#Fix the date
pm25$Date <- mdy(pm25$Date)

```


## Explore properties of the dataframe
-   How many different sites are included in the data set?
-   What is the mean daily PM2.5 concentration?
-   In what units of are PM 2.5 concentrations reported?

```{r Explore the data}
# Compute the number of different sites
length(unique(pm25$Site.ID))

# Compute the mean and median PM 2.5 concentration
mean(pm25$Daily.Mean.PM2.5.Concentration, na.rm = TRUE)
median(pm25$Daily.Mean.PM2.5.Concentration, na.rm = TRUE)

# What are the units
unique(pm25$UNITS)

# What is the date range (min and max dates)
min(pm25$Date, na.rm = TRUE)
max(pm25$Date, na.rm = TRUE)

```


## Plot distribution of values

-   Number of samples using each different source 
-   Distribution of daily mean concentration values
-   Distribution of daily mean concentration, by source*
  *   There are multiple ways to visualize this...

```{r Visualize the data}
#Number of samples using each different source 
ggplot(pm25, aes(x = Source)) +
  geom_bar() +
  labs(
    title = "Number of samples by source",
    x = "Source",
    y = "Count"
  )

#Distribution of daily mean concentration values
ggplot(pm25, aes(x = Daily.Mean.PM2.5.Concentration)) +
  geom_histogram(bins = 30, color = "black") +
  labs(
    title = "Distribution of daily mean PM2.5 concentration",
    x = "Daily mean PM2.5 concentration",
    y = "Frequency"
  )


#Distribution of daily mean concentration, by source
ggplot(pm25, aes(
  x = Source,
  y = Daily.Mean.PM2.5.Concentration
)) +
  geom_boxplot() +
  labs(
    title = "PM2.5 concentration by source",
    x = "Source",
    y = "Daily mean PM2.5 concentration"
  )



```

